---
title: "Network Analytics (2623) — Group Assignment 2"
output:
  html_document:
    toc: true
    toc_float: true
    theme: flatly
    code_folding: show
---

# INSTRUCTIONS

-   This is a group assignment.
-   Submit your answer digitally as two files through Moodle:
    -   An R markdown file (extension **Rmd**). Use the template provided
        to you and provide your answers (both code and text) below each
        question.
    -   An **HTML** file "knitted" by RStudio including all the results and
        plots. More details on how to create these files will be provided
        in class on week 3.
-   Follow the Style Guide (available on Moodle). You can be
    penalized on up to 20% in each question for which you do not follow
    the Style Guide.
-   Questions regarding the assignment should be posted <span class="underline">exclusively</span> on
    the respective discussion forum on Moodle.
-   **Deadline:** Sunday, 3 May 2026, 23:59 (Lisbon time)

<span class="underline">**Warning:**</span> The detection of <span class="underline">any form of plagiarism</span> in your work
means the assignment will be graded with <span class="underline">ZERO points</span>.

# Movie Networks

We are interested in assessing what are the most important movies in
the decade 2010-2019. We will use different strategies to do so.
First, we will load and prepare the data.

## Load and prepare the data

The first step is to load and prepare the movie data. The following
instructions perform some routine data preparation operations. Each
set of instructions is preceded by a comment explaining the procedure.
Run the code below and try to understand each line of code as you
might need to perform some changes.

```{r}
library(data.table)     # Run once per session
library(ggplot2)        # Run once per session

# Load data from file 20200120-imdb_movie_actor.csv (do not forget to
# change your working directory to the folder containing the
# 20200120-imdb_movie_actor.csv file)
dt.movie.actor <- fread("20200120-imdb_movie_actor.csv") 

# Count in how many movies each actor has participated and how many 
# principal actor each movie has
dt.movie.actor[, n_movies := .N, by=actor]
dt.movie.actor[, n_actors := .N, by=list(movie, year)]

# Remove entries in which actors have no name 
dt.movie.actor <- dt.movie.actor[!(actor == "")]

# Save dt.movie.actor. Next time you can simply call the load function (below)
save(dt.movie.actor, file="imdb_movie_actor.RData") 
```

Load the data that you prepared using the instructions below. As mentioned in the
comments, you can start from this line if you have previously saved these data.

```{r}
# Load previously saved dt.movie.actor. You can
# start in this line if you have previously saved these data.
load("imdb_movie_actor.RData") 
```

## Questions (`data.table`) `[7 points]`

This set of questions require that you know how to manipulate a
`data.table`. Answer each of the following questions below. Include
all the code you created/used in your answer.

1.  What is the total amount of movies in the `dt.movie.actor` dataset?
    `[1 point]`

```{r}
# <answer here>
```

2.  List the actors from the movie `"Fight Club (1999)"`. List
    the actors from the movie `"Se7en (1995)"`. `[1 point]`

```{r}
# <answer here>
```

3.  Which actors participated on both movies? Hint: The function
    `intersect` calculates the intersection of two sets. `[1 point]`

```{r}
# <answer here>
```

4.  In which movies did Brad Pitt (b.1963) and George Clooney (b.1961) star together?
    `[1 point]`

```{r}
# <answer here>
```

5.  Create a table that shows the number of movies released per year.
    This table should include three columns: `year`, `n_movies`, and
    `csum_n_movies`. The first column should contain the year, the
    second the number of movies in that year, and the third, the number
    of movies released since the first year in the data and up to the
    year in that line. Tip: Use the function `cumsum` and check if the
    amount in the last year is the same as the total number of movies
    in question 1. `[1 point]`

```{r}
# <answer here>
```

6.  Which actor/actress has starred in the most movies across all data?
    After (and including) 2000, which year has the most movie
    participations by a single actor/actress? Who is that
    actor/actress? What do these two actors/actresses have in common?
    `[1 point]`

```{r}
# <answer here>
```

7.  Consider only the 10% most popular movies (by votes) in the decade 2010-2019.
    List the top 10 actors that starred in the most movies in the decade.
    Which year(s) has/have the most movie participations by a single actor?
    Hint: you can use the function `quantile` to find how many votes
    does the movie in percentile 90 have. `[1 point]`

```{r}
# <answer here>
```

## Questions (`ggplot2`) `[3 points]`

1.  Plot a histogram with the number of movies per year. Which patterns
    do you observe? Is there anything strange? `[1 point]`

```{r}
dt.movies.per.year <- dt.movie.actor[, .(n_movies = uniqueN(movie)), by = year]

ggplot(dt.movies.per.year, aes(x = year, y = n_movies)) +
  geom_bar(stat = "identity")
```
Movie production grew steadily from the early 1900s, with a sharp acceleration from the 1990s onwards, reflecting the globalisation of film production and the rise of digital filmmaking. There is a noticeable dip around the 1930s-40s, which can likely be explained by the Great Depression and World War II, both of which severely impacted the film industry. After the war, production recovered and grew gradually through the 1960s-80s before exploding in the 1990s and 2000s. One strange observation is that the last bar (2019) appears slightly lower than 2017 and 2018, which is unexpected given the general upward trend. This is most likely because the dataset was collected in January 2020, meaning not all 2019 movies had been added to IMDb yet at the time of collection.

2.  Plot a histogram that represents the distribution of number of IMDb
    votes per movie. The x-axis should represent the number of votes
    and the y-axis should represent how many movies have x number of
    votes. Which patterns do you observe? `[1 point]`

```{r}
dt.votes <- unique(dt.movie.actor[votes > 0, .(movie, votes)])

ggplot(dt.votes, aes(x = votes)) +
  geom_histogram() +
  scale_x_log10()
```
The distribution of IMDb votes per movie is extremely right-skewed. The vast majority of movies in the dataset have received very few votes while a tiny fraction of well-known blockbusters have accumulated hundreds of thousands or even millions of votes. This makes intuitive sense, as most movies in the dataset are obscure or low-budget productions that very few people have watched or rated. The log scale makes the shape much clearer, revealing that the distribution peaks before 100 votes and then tapers off gradually towards the right.

3.  Plot a histogram that represents the distribution of the number of
    actors per movie. The x-axis should represent the number of actors
    and the y-axis should represent how many movies have x number of actors.
    Describe your findings. `[1 point]`

```{r}
dt.actors.per.movie <- unique(dt.movie.actor[, .(movie, n_actors)])

ggplot(dt.actors.per.movie, aes(x = n_actors)) +
  geom_histogram(binwidth = 1)
```
The vast majority of movies have exactly 4 actors listed, which is likely a structural feature of how IMDb records principal cast. Very few movies have more than 8 or fewer than 3 actors.

## Questions (`igraph`) `[10 points]`

1.  From this question onwards, and until the end of the assignment,
    focus only on <span class="underline">the actors that participated on the top 50 most
    popular movies from the 2010-2019 decade (by number of votes).</span>
    Load the `igraph` package and create a bipartite graph in which the
    edges correspond to actors' participation in movies. How many movie
    participations exist? `[1 point]`

```{r}
# Filter to actors from top 50 most popular movies of 2010-2019 decade
top.movies.2010.2019 <- dt.movie.actor[year >= 2010 & year <= 2019, 
    .N, by = list(movie, votes)
    ][order(-votes)][1:50, movie]

actors.in.top <- dt.movie.actor[movie %in% top.movies.2010.2019, 
                                unique(actor)]

# Create bipartite graph with actors and movies
edges <- dt.movie.actor[movie %in% top.movies.2010.2019, 
                       list(actor, movie)]

# Create bipartite graph
g.bipartite <- graph_from_data_frame(edges, directed = FALSE)

# Mark bipartite attribute
V(g.bipartite)$type <- !V(g.bipartite)$name %in% unique(edges$movie)


# Summary and number of movie participations
summary(g.bipartite)
ecount(g.bipartite) 
```
There are 200 movie participations.


2.  Create a graph in which two movies are connected to each other if
    they have <span class="underline">at least one actor in common</span>. Calculate the <span class="underline">degree
    centrality</span> for each of the movies, and remove movies with no
    connections to other movies. <span class="underline">Hint:</span> the function
    `induced.subgraph` allows the creation of graphs with only a subset
    of the vertices. Calculate the following additional centrality
    measures for each of these movies: `[2 points]`
    -   Closeness centrality
    -   Betweenness centrality
    -   Eigenvector centrality

```{r}
# Create vertex list with type attribute
all.actors <- dt.movie.actor[movie %in% top.movies.2010.2019, list(name = unique(actor), type = TRUE)]
all.movies <- dt.movie.actor[movie %in% top.movies.2010.2019, list(name = unique(movie), type = FALSE)]
all.vertices <- rbind(all.actors, all.movies)

# Build bipartite graph
g <- graph.data.frame(dt.movie.actor[movie %in% top.movies.2010.2019,  
                                     list(movie, actor)
                                     ], directed = FALSE, vertices = all.vertices)

# Project to movie space 
g.movies <- bipartite.projection(g)$proj1
summary(g.movies)

# Calculate degree centrality
V(g.movies)$degree <- degree(g.movies)

# Remove movies with no connections (degree == 0)
g.movies.connected <- induced.subgraph(g.movies, vids = V(g.movies)[V(g.movies)$degree > 0])
summary(g.movies.connected)

# Calculate all centrality measures
V(g.movies.connected)$degree      <- degree(g.movies.connected)
V(g.movies.connected)$closeness   <- closeness(g.movies.connected)
V(g.movies.connected)$betweenness <- betweenness(g.movies.connected)
V(g.movies.connected)$evcent      <- evcent(g.movies.connected)$vector

# Convert to data.table for analysis
dt.g.movies <- data.table(get.data.frame(g.movies.connected, "vertices"))
head(dt.g.movies[order(-betweenness)], 10)
```

3.  For each centrality measure, list the top 20 movies with highest
    centrality. How do you interpret the outcomes? In addition, report the
    exact betweenness centrality score of the **top-ranked movie by
    betweenness** to four decimal places. `[2 points]`

```{r}
# Top 20 by each centrality measure
head(dt.g.movies[order(-degree)], 20)
head(dt.g.movies[order(-closeness)], 20)
head(dt.g.movies[order(-betweenness)], 20)
head(dt.g.movies[order(-evcent)], 20)

# Report exact betweenness of top-ranked movie
dt.g.movies[order(-betweenness)][1, .(name, betweenness = round(betweenness, 4))]
```
Interpretation of Centrality Measures:
Degree Centrality: 
The Avengers series dominates with the highest degree of centrality, with each film having 11 connections. This indicates that these films share the most actors with other movies in the top 50, making them highly connected hubs in the actor-sharing network.
Closeness Centrality: 
X-Men: First Class (2011) has the highest closeness centrality score (0.0556), followed by other films with large casts, such as Silver Linings Playbook (2012) and X-Men: Days of Future Past (2014). On average, these movies are closer to the other movies, suggesting that they occupy more central positions in the network's overall structure.
Betweenness Centrality: 
Shutter Island (2010) takes the top spot with a score of 144.4286, followed by Now You See Me (2013) with 84.4286 and The Revenant (2015) with 63.50. These films act as crucial bridges, connecting different clusters of films through shared actors. Shutter Island is especially important in connecting parts of the network that would otherwise be disconnected.
Eigenvector Centrality: 
The three Avengers films received perfect scores of 1.0, indicating that they are connected to other highly connected movies. Captain America: Civil War (2016) and The Avengers (2012) also received high scores (0.78), demonstrating their connection to significant nodes in the network.

Key Insight: Different films are central for different reasons. The Avengers films are highly connected and influential, while Shutter Island acts as a critical bridge between different movie clusters. Films with high closeness centrality, such as X-Men: First Class, occupy more balanced positions within the network.

The exact betweenness centrality of the top-ranked movie is: Shutter Island (2010): 144.4286



4.  Calculate the average clustering coefficient for the movies
    network. *Hint: use `transitivity(g, type = "global")` to calculate
    the average clustering coefficient. For movie co-occurrence networks
    of this type, values typically fall in the range 0.75--0.85.*
    `[1 point]`

```{r}
# <answer here>
```

5.  Choose one movie you like and plot the movie, their direct
    neighbors and the links among them. What is the clustering
    coefficient of this movie? Which is the actor with most
    participations among these (neighbor) movies, but not having
    participated in the movie itself? *Note: in your written answer,
    explicitly state the ratio of your chosen movie's clustering
    coefficient to the network average you computed in Question 4
    (i.e. report both numbers and their ratio to two decimal places).*
    `[2 points]`

```{r}
# <answer here>
```

6.  Plot the degree distribution of the movies. How do you compare it
    with the degree distribution of a random graph with the same number
    of nodes and edges? What can be plausible explanations for the
    observed differences? *In your answer, name one movie that ranks in
    the **top 5 by betweenness centrality** but does **not** appear in
    the **top 10 by degree centrality**. Provide its exact betweenness
    centrality value (to four decimal places) and explain why such a
    movie might act as a critical bridge despite having relatively few
    direct connections.* `[2 points]`

```{r}
# <answer here>
```
