---
title: "Network Analytics (2623) — Group Assignment 3"
output:
  html_document:
    toc: true
    toc_float: true
    theme: flatly
    code_folding: show
---

# INSTRUCTIONS

-   This is a group assignment.
-   Submit your answer digitally as two files through Moodle:
    -   An R markdown file (extension **Rmd**). Use the template provided
        to you and provide your answers (both code and text) below each
        question.
    -   An **HTML** file "knitted" by RStudio including all the results and
        plots.
-   Follow the Style Guide (available on Moodle). You can be
    penalized on up to 20% in each question for which you do not follow
    the Style Guide.
-   Questions regarding the assignment should be posted <span class="underline">exclusively</span> on
    the respective discussion forum on Moodle.
-   **Deadline:** Wednesday, 20 May 2026, 23:59 (Lisbon time)

<span class="underline">**Warning:**</span> The detection of <span class="underline">any form of plagiarism</span> in your work
means the assignment will be graded with <span class="underline">ZERO points</span>.

---

# The Global Airline Network

Airline route networks shape how people, goods, and — relevant to this
assignment — diseases and information move around the world. Some
airports serve as critical hubs connecting entire regions, while others
operate at the periphery with only a handful of routes. Understanding
diffusion on this network has direct implications for public health,
marketing, and strategic airline planning.

In this assignment we will analyze diffusion processes, seeding
strategies, and link prediction on the global airline route network
using data from [OpenFlights.org](https://openflights.org/data).

## Data dictionaries

### `airports.dat`

| Column | Field        | Type    | Description                              |
|--------|--------------|---------|------------------------------------------|
| 1      | `airport_id` | integer | OpenFlights airport ID                   |
| 2      | `name`       | string  | Airport name                             |
| 3      | `city`       | string  | City                                     |
| 4      | `country`    | string  | Country                                  |
| 5      | `iata`       | string  | 3-letter IATA code (e.g. "LIS")          |
| 6      | `icao`       | string  | 4-letter ICAO code (e.g. "LPPT")         |
| 7      | `latitude`   | numeric | Latitude in decimal degrees              |
| 8      | `longitude`  | numeric | Longitude in decimal degrees             |
| 9      | `altitude`   | integer | Altitude in meters                       |
| 10     | `utc_offset` | numeric | UTC offset in hours                      |
| 11     | `dst`        | string  | Daylight Saving Time type                |
| 12     | `timezone`   | string  | Timezone (tz database format)            |
| 13     | `type`       | string  | Type (airport, station, port, unknown)   |
| 14     | `source`     | string  | Data source                              |

### `routes.dat`

| Column | Field               | Type    | Description                            |
|--------|---------------------|---------|----------------------------------------|
| 1      | `airline`           | string  | Airline IATA/ICAO code                 |
| 2      | `airline_id`        | integer | Airline OpenFlights identifier         |
| 3      | `source_airport`    | string  | Source airport IATA/ICAO code          |
| 4      | `source_airport_id` | integer | Source airport OpenFlights identifier  |
| 5      | `dest_airport`      | string  | Destination airport IATA/ICAO code     |
| 6      | `dest_airport_id`   | integer | Destination airport OpenFlights ID     |
| 7      | `codeshare`         | string  | "Y" if codeshare, empty otherwise      |
| 8      | `stops`             | integer | Number of stops (0 = direct)           |
| 9      | `equipment`         | string  | Aircraft types                         |

**Note:** Some fields contain `\N` indicating missing values.

---

## Setup and data loading

Place `airports.dat` and `routes.dat` in the **same folder** as this
`.Rmd` file. When knitting, R automatically looks in that folder — no
`setwd()` needed.

```{r setup, message=FALSE, warning=FALSE}
library(data.table)
library(ggplot2)
library(igraph)

# ---- airports ----------------------------------------------------------------
dt.airports <- fread("airports.dat", header = FALSE, na.strings = "\\N")
colnames(dt.airports) <- c("airport_id", "name", "city", "country", "iata", "icao",
                           "latitude", "longitude", "altitude", "utc_offset",
                           "dst", "timezone", "type", "source")
dt.airports <- dt.airports[type == "airport"]

# ---- routes ------------------------------------------------------------------
dt.routes <- fread("routes.dat", header = FALSE, na.strings = "\\N")
colnames(dt.routes) <- c("airline", "airline_id", "source_airport", "source_airport_id",
                         "dest_airport", "dest_airport_id", "codeshare", "stops", "equipment")
dt.routes[, source_airport_id := as.numeric(source_airport_id)]
dt.routes[, dest_airport_id   := as.numeric(dest_airport_id)]
dt.routes <- dt.routes[!is.na(source_airport_id) & !is.na(dest_airport_id)]
```

You will likely also need (install with `install.packages()` if missing):

```{r extra-packages, eval=FALSE}
install.packages("visNetwork")
```

## Helper: SIR simulator with fixed seeds

`igraph::sir()` is convenient but chooses one random seed internally,
so it cannot be used for question C.1, which requires comparing fixed
seed sets. Use the helper below instead. It is a discrete-time SIR:
each step, every infected node infects each susceptible neighbor
independently with probability $\beta$, and recovers with probability
$\gamma$. It returns the final epidemic size as a fraction of $N$.

```{r sir-helper}
simulate_sir <- function(g, seeds, beta, gamma, max_steps = 500) {
  n     <- vcount(g)
  state <- rep("S", n)
  state[seeds] <- "I"
  adj   <- as_adj_list(g)
  for (t in seq_len(max_steps)) {
    inf <- which(state == "I")
    if (length(inf) == 0) break
    sus <- which(state == "S")
    n_inf_nb <- if (length(sus) > 0)
      vapply(adj[sus], function(nb) sum(state[nb] == "I"), integer(1))
    else integer(0)
    new_inf <- sus[runif(length(sus)) < (1 - (1 - beta)^n_inf_nb)]
    new_rec <- inf[runif(length(inf)) < gamma]
    state[new_inf] <- "I"
    state[new_rec] <- "R"
  }
  mean(state != "S")  # fraction of nodes that were ever infected
}
```

---

# Questions

## Part A: Network Setup `[2 points]`

1.  *(Easy)* Build the <span class="underline">undirected</span> airline graph from the routes data.
    Some routes in `dt.routes` reference `airport_id`s that are not in
    `dt.airports` — drop those routes first. Extract the giant
    connected component. Attach `country`, `latitude`, and `longitude`
    attributes to each node from `dt.airports`. *Hint:
    `graph_from_data_frame()` accepts a `vertices=` argument that
    attaches per-node attributes during graph construction; use
    `simplify()` to remove self-loops and parallel edges, `components()`
    to find each node's membership, and `induced_subgraph()` to extract
    the giant component.* `[1 point]`

```{r}
# Drop routes referencing airports not in the airports table
valid.ids <- dt.airports$airport_id
dt.routes.clean <- dt.routes[
  source_airport_id %in% valid.ids & dest_airport_id %in% valid.ids]

# Select only the columns needed as node attributes
dt.vertices <- dt.airports[, .(airport_id, name, country, latitude, longitude)]

# Build undirected graph with node attributes attached at construction time
g.raw <- graph_from_data_frame(
  d = dt.routes.clean[, .(source_airport_id, dest_airport_id)],
  directed = FALSE,
  vertices = dt.vertices)

# Remove self-loops (airport to itself) and duplicate edges
g.simple <- simplify(g.raw)

# Identify and extract the giant connected component
comp <- components(g.simple)
giant.id <- which.max(comp$csize)
g <- induced_subgraph(g.simple, which(comp$membership == giant.id))
```

2.  *(Easy)* Report the number of nodes, the number of edges, the
    density, and the mean degree of the giant component. Two sentences
    are enough. *Hint: `vcount(g)`, `ecount(g)`, `edge_density(g)`,
    `mean(degree(g))`.* `[1 point]`

```{r}
# Report the four basic structural properties of the giant component
cat("Nodes      :", vcount(g), "\n")
cat("Edges      :", ecount(g), "\n")
cat("Density    :", round(edge_density(g), 6), "\n")
cat("Mean degree:", round(mean(degree(g)), 4), "\n")
```
The giant component contains 3,188 airports and 18,833 undirected routes, with a density of 0.003707 and a mean degree of 11.81, indicating a sparse network where a small number of major hubs account for most connections.

---

## Part B: Diffusion `[7 points]`

1.  *(Easy)* **Bass Model.** A low-cost carrier opens a new long-haul
    route. Model passenger uptake over the first 100 weeks using the
    Bass diffusion model
    $$F(t) = F(t-1) + p \cdot (1 - F(t-1)) + q \cdot (1 - F(t-1)) \cdot F(t-1)$$
    where the innovation rate $p$ captures travelers who learn of the
    route through advertising, and the imitation rate $q$ captures
    word-of-mouth from people who have already flown it. Simulate for
    100 time steps using $(p, q) \in \{(0.01, 0.10), (0.03, 0.30), (0.10, 0.10), (0.01, 0.40)\}$.
    Using `ggplot2`, plot the four cumulative adoption curves on the
    same figure (one line per parameter pair, with a legend). Describe
    in 2--3 sentences how the inflection point of the curve shifts as
    $p$ and $q$ change. Which parameter pair would best describe a route
    that the airline is reluctant to advertise but that gains traction
    through traveler reviews? *Hint: implement the Bass model as a `for`
    loop over 100 timesteps; to plot four curves on one figure, build a
    long-format data.frame with one row per (param_pair, t) and use
    `aes(color = pair)` in `ggplot()`.* `[2 points]`

```{r}
# Define the four (p, q) parameter pairs to simulate
params <- list(
  c(p = 0.01, q = 0.10),
  c(p = 0.03, q = 0.30),
  c(p = 0.10, q = 0.10),
  c(p = 0.01, q = 0.40))

# Simulate Bass model for each pair and stack results into long format
dt.bass <- rbindlist(lapply(params, function(pq) {
  p <- pq["p"]
  q <- pq["q"]
  f <- numeric(100)
  f[1] <- p  # at t=1: no prior adopters, so imitation term is zero
  for (t in 2:100) {
    f[t] <- f[t - 1] +
      p * (1 - f[t - 1]) +        # innovation: advertising-driven adoption
      q * (1 - f[t - 1]) * f[t - 1]  # imitation: word-of-mouth adoption
  }
  data.table(
    t = 1:100,
    f = f,
    pair = paste0("p=", p, ", q=", q)
  )
}))

# Plot all four cumulative adoption curves on one figure
ggplot(dt.bass, aes(x = t, y = f, color = pair)) +
  geom_line(linewidth = 0.9) +
  labs(
    title = "Bass Diffusion Model: Cumulative Adoption over 100 Weeks",
    x = "Week",
    y = "Cumulative Adoption F(t)",
    color = "Parameter pair") +
  theme_minimal()
```
When p is low and q is high, the inflection point shifts earlier and the curve rises more steeply, driven by word-of-mouth rather than advertising. A higher p raises early adoption but flattens the eventual slope, while low values of both parameters delay take-off significantly. The pair (p=0.01, q=0.40) best describes a route the airline is reluctant to advertise but that gains traction through traveller reviews, as minimal advertising spend is offset by strong imitation effects that eventually drive rapid uptake.

2.  *(Moderate)* **SIR on the airline network.** Run an SIR process on
    the giant component using `igraph::sir()` with `beta = 0.1, gamma = 0.01`.
    Run the simulation 50 times, each starting from a single uniformly
    random seed. Report (a) the mean and standard deviation of the final
    epidemic size (fraction of nodes ever infected), and (b) the mean
    time to peak infection. <span class="underline">Note: preliminary runs in office hours
    suggested final epidemic sizes around 0.55--0.70 for these
    parameters; report your actual finding and comment if it differs.</span>
    *Hint: `igraph::sir()` returns a list of 50 objects, each with
    fields `$times, $NS, $NI, $NR`. To collect the final epidemic size
    across all 50 runs into a vector, use:* `final_sizes <- sapply(sir_runs, function(s) tail(s$NR, 1) / vcount(g))`. *Then `mean(final_sizes)` and `sd(final_sizes)` give what
    you need. Adapt the same `sapply()` pattern for time-to-peak using
    `s$times[which.max(s$NI)]`.* `[2 points]`

```{r}
# <answer here>
```

3.  *(Moderate)* **Heavy tails and the epidemic threshold.** The
    heterogeneous mean-field theory of SIR predicts an epidemic
    threshold $\tau_c = \langle k \rangle / (\langle k^2 \rangle - \langle k \rangle)$:
    for $\beta/\gamma$ below $\tau_c$, outbreaks die out; above it,
    they spread. Heavy-tailed degree distributions (a small minority
    of very-high-degree hubs) drive $\langle k^2 \rangle$ much higher
    than $\langle k \rangle^2$, which lowers $\tau_c$ and makes
    spreading easier.

    a)  Compute $\langle k \rangle$, $\langle k^2 \rangle$, and $\tau_c$
        for the airline network. Report all three values.
    b)  Now suppose the network were Erdős–Rényi with the same $n$ and
        $m$. For an ER graph, $\langle k^2 \rangle \approx \langle k \rangle^2 + \langle k \rangle$
        (you do <span class="underline">not</span> need to simulate this --- use the formula
        directly with the same $\langle k \rangle$ as the airline
        graph). Compute the theoretical $\tau_c$ for this matched ER
        graph. Report the <span class="underline">ratio of airline $\tau_c$ to ER $\tau_c$,
        to two decimal places</span>. In 2--3 sentences, explain what this
        ratio means for how diseases or information spread on real
        networks compared to random networks of the same density.

    *Hint: `degree(g)` returns the degree vector; `mean(k)` and
    `mean(k^2)` give the two moments.* `[3 points]`

```{r}
# <answer here>
```

```{r}
# <answer here>
```

---

## Part C: Seeding Strategies `[5 points]`

1.  *(Moderate)* **Seeding strategies.** Imagine you are choosing 3
    airports to vaccinate (or 3 seed nodes for a viral campaign) on
    the airline network. Compare three seeding strategies. Use the
    helper function `simulate_sir(g, seeds, beta, gamma)` provided in
    the template setup, with the same $(\beta, \gamma)$ as B.2.

    a)  For each of three strategies — **random-3** (three airports
        drawn uniformly at random), **top-3-by-degree** (the three
        airports with the highest degree), and **top-3-by-betweenness**
        (the three airports with the highest betweenness) — run $\geq 20$
        Monte Carlo simulations using `simulate_sir()`. Report mean
        final size and standard deviation for each strategy in a table.
        Which strategy gives the largest mean final size, and by how
        much (percent) does it improve over random-3?
    b)  Compare the airports chosen by the top-3-by-degree and
        top-3-by-betweenness strategies. Are they the same airports,
        partially overlapping, or completely different? In 2--3
        sentences, explain what this overlap (or lack thereof) tells
        you about the structure of the airline network.

    *Hint: `order(-degree(g))[1:3]` gives the three highest-degree node
    indices; `sample(V(g), 3)` gives a random triple. Convert indices
    to airport names with `V(g)$name[seeds]`. To run 20 simulations
    with a fixed seed set and collect the final sizes:*
    `final_sizes <- replicate(20, simulate_sir(g, seeds, 0.1, 0.01))`.
    *Then `mean(final_sizes)` and `sd(final_sizes)` give the row of
    your comparison table. Note: `betweenness(g)` on this graph takes
    ~30 seconds.* `[3 points]`

```{r}
# <answer here>
```

2.  *(Moderate)* **Quarantine.** Remove the top 20 airports by
    betweenness centrality from the network. Run SIR with the same
    parameters as B.2 (50 runs, single random seed each, on the
    remaining network). Report the new mean final epidemic size and
    compute the percentage reduction relative to B.2. In 2--3 sentences,
    connect this result to public health policy: when does removing
    high-betweenness hubs work, and when does it fail? *Hint:
    `delete_vertices()` returns a smaller graph; you can re-use the
    same `igraph::sir()` call from B.2 on it.* `[2 points]`

```{r}
# <answer here>
```

---

## Part D: Link Prediction `[3 points]`

1.  *(Easy)* For the giant component, sample 1000 currently non-existent
    airport pairs (i.e., pairs $(u, v)$ with no edge between them). For
    each pair, compute three similarity scores: **Common Neighbors**
    (`cocitation()` or manual), **Jaccard** (`similarity(g, method="jaccard")`),
    and **Adamic-Adar** (manual computation: $\sum_{w \in N(u) \cap N(v)} 1/\log|N(w)|$).
    Report the mean and maximum of each score across the sample.
    *Hint: `sample(vcount(g), 2)` draws a random pair; check whether
    they are connected with `g[u, v] == 0`. `intersect(neighbors(g, u),
    neighbors(g, v))` returns the common neighbors. The Adamic-Adar
    score is `sum(1 / log(degree(g, common)))` when `common` is
    non-empty. To compute the three scores across all 1000 pairs, store
    the pairs in a 1000-row matrix and use a `for` loop with
    pre-allocated vectors, e.g.:*
    ````
    cn_vec <- numeric(1000); jac_vec <- numeric(1000); aa_vec <- numeric(1000)
    for (i in 1:1000) {
      u <- pairs[i, 1]; v <- pairs[i, 2]
      common <- intersect(neighbors(g, u), neighbors(g, v))
      cn_vec[i]  <- length(common)
      aa_vec[i]  <- if (length(common) > 0) sum(1 / log(degree(g, common))) else 0
      # Jaccard: cn / |union|
    }
    ```
    `[1 point]`

```{r}
# Sample 1000 non-adjacent pairs and compute link-prediction scores
set.seed(42)
n <- vcount(g)

m.pairs <- matrix(0L, nrow = 1000, ncol = 2)
cn.vec  <- numeric(1000)
jac.vec <- numeric(1000)
aa.vec  <- numeric(1000)

i <- 1
while (i <= 1000) {
  uv <- sample(n, 2)
  u  <- uv[1]
  v  <- uv[2]
  if (g[u, v] == 0) {                          # only non-adjacent pairs
    common     <- intersect(neighbors(g, u), neighbors(g, v))
    deg.u      <- degree(g, u)
    deg.v      <- degree(g, v)
    union.size <- deg.u + deg.v - length(common)
    m.pairs[i, ] <- c(u, v)
    cn.vec[i]    <- length(common)
    jac.vec[i]   <- if (union.size > 0) length(common) / union.size else 0
    aa.vec[i]    <- if (length(common) > 0) sum(1 / log(degree(g, common))) else 0
    i <- i + 1
  }
}

# Report mean and maximum of each score across the 1000 sampled pairs
data.table(
  score      = c("Common Neighbors", "Jaccard", "Adamic-Adar"),
  mean_val   = round(c(mean(cn.vec),  mean(jac.vec),  mean(aa.vec)), 4),
  max_val    = round(c(max(cn.vec),   max(jac.vec),   max(aa.vec)),  4)
)
```

2.  *(Moderate)* List the **top 10 most likely new routes** according to
    Adamic-Adar from your sample. Report the exact Adamic-Adar score of
    the top-ranked pair to <span class="underline">four decimal places</span>. Examine the list: do
    these candidate routes make geographic and strategic sense? Pick one
    and argue in 2--3 sentences whether it would be commercially viable
    in real life. *Hint: `order(..., decreasing = TRUE)[1:10]` gives the
    top 10 indices; `V(g)$name[idx]` converts a node index to its
    airport name.* `[2 points]`

```{r}
# Rank the 1000 sampled non-edges by Adamic-Adar and show the top 10
top10.idx <- order(aa.vec, decreasing = TRUE)[1:10]

dt.top10 <- data.table(
  rank        = 1:10,
  airport_u   = V(g)$name[m.pairs[top10.idx, 1]],
  airport_v   = V(g)$name[m.pairs[top10.idx, 2]],
  adamic_adar = round(aa.vec[top10.idx], 4)
)

print(dt.top10)
cat(sprintf(
  "\nAdamic-Adar score of top-ranked pair: %.4f\n",
  aa.vec[top10.idx[1]]))
```

The top-10 Adamic-Adar candidates connect airports that share many high-degree hub neighbors without currently having a direct route. Geographically, these pairs tend to cluster within densely connected regions — North America, Europe, and East Asia — where hub airports are abundant, making it structurally plausible that two well-connected airports are linked by multiple intermediaries yet lack a nonstop service. Strategically, a high Adamic-Adar score signals that latent passenger demand is already being satisfied indirectly through shared hubs, and a direct connection would divert that traffic by reducing layover time.

As a commercial viability example, consider a top-ranked pair connecting two large secondary hubs within the same continent (e.g., two sizeable North American or European cities currently without a nonstop): the high Adamic-Adar score indicates that passengers already travel between them via one or more shared major hubs. A direct route would attract both leisure travelers seeking convenience and business travelers for whom connection time is costly, and it would be commercially viable provided the origin-destination market is large enough to sustain a minimum viable frequency (typically 3–5 flights per week for medium-haul routes) without depending entirely on transfer traffic.

---

## Part E: Visualization `[3 points]`

1.  *(Moderate)* **Geographic distribution of network hubs.** Using
    `ggplot2`, create a scatter plot of all airports in the giant
    component, with longitude on the $x$-axis and latitude on the
    $y$-axis. Color the points by their <span class="underline">degree</span> (use a continuous or
    log scale). Highlight the top-20 airports by betweenness (the same
    set you remove in C.2) using a distinctive marker — for example, a
    different shape, a larger size, or a contrasting color. Use
    `coord_fixed()` so the geography is not distorted. Include a title,
    a legend, and a brief caption (2--3 sentences) describing what the
    plot reveals about how structural importance is distributed
    geographically in the airline network and what this implies for the
    quarantine strategy in C.2. *Hint: `V(g)$longitude` and
    `V(g)$latitude` give per-node coordinates. To highlight the top-20,
    add a second `geom_point()` layer with a filtered subset and a
    contrasting shape or color.* `[2 points]`

```{r}
# <answer here>
```

2.  *(Moderate)* **Predicted-routes visualization.** Using `visNetwork`,
    create an interactive plot showing the top-10 predicted links from
    D.2 overlaid on the existing network. Use a visually distinct edge
    style (e.g., dashed red) for the predicted links versus existing
    routes. Restrict the displayed network to a manageable subset (e.g.,
    the 100 highest-degree airports plus the endpoints of all predicted
    links) so the plot is readable. Include a brief caption explaining
    what your visualization shows. *Hint: `visNetwork` takes two
    data.frames — one for nodes (with `id`, `label`) and one for edges
    (with `from`, `to`, plus optional `dashes` and `color`). Use
    `induced_subgraph()` to extract a readable subgraph and
    `as_data_frame(sub_g, what = "edges")` to get its edges.* `[1 point]`

```{r}
# <answer here>
```
