---
title: "Network Analytics - Group Assignment 2"
author: "Rollandin Alessandro, Titze Konstantin, Haussmann Maximilian, Ciorciolini Francesco, Borella Riccardo"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE)

library(data.table)
library(ggplot2)
library(igraph)

dt.movie.actor <- fread("20200120-imdb_movie_actor.csv")
dt.movie.actor[, n_movies := .N, by = actor]
dt.movie.actor[, n_actors := .N, by = list(movie, year)]
dt.movie.actor <- dt.movie.actor[!(actor == "")]
save(dt.movie.actor, file = "imdb_movie_actor.RData")

load("imdb_movie_actor.RData")
```

## Part 1: `data.table`
### Question 1
```{r}
num_movies <- dt.movie.actor[, uniqueN(movie)]
print(paste("Number of unique movies:", num_movies))
```

### Question 2
```{r}
actors_fight_club <- dt.movie.actor[movie == "Fight Club (1999)", unique(actor)]
print(paste("Actors in Fight Club (1999):", paste(actors_fight_club, collapse = ", ")))

actors_se7en <- dt.movie.actor[movie == "Se7en (1995)", unique(actor)]
print(paste("Actors in Se7en (1995):", paste(actors_se7en, collapse = ", ")))
```

### Question 3
```{r}
actor_both_movies <- intersect(actors_fight_club, actors_se7en)
print(paste("Actor(s) in both movies:", paste(actor_both_movies, collapse = ", ")))
```

### Question 4
```{r}
movie_brad_pitt_george_clooney <- dt.movie.actor[actor %in% c("Brad Pitt (b.1963)", "George Clooney (b.1961)"), unique(movie)]
print(paste("Movie(s) with Brad Pitt and George Clooney:", paste(movie_brad_pitt_george_clooney, collapse = ", ")))
```

### Question 5
```{r}
num_movies_per_year <- dt.movie.actor[, .(n_movies = uniqueN(movie)), by = year][order(year)][, csum_n_movies := cumsum(n_movies)]
print("Number of movies per year:")
print(num_movies_per_year)
```

### Question 6
Add what they have in common
```{r}
top_actor_all <- dt.movie.actor[, .(n = .N), by = actor][order(-n)][1]
print(paste("Top actor overall:", top_actor_all$actor))

top_year_actor <- dt.movie.actor[year >= 2000, .(n = .N), by = .(year, actor)][order(-n)][1]
print(paste("Year with the most movies participations by a single actor/actress:", top_year_actor$year))
print(paste("Top actor in", top_year_actor$year, ":", top_year_actor$actor))

#check to see if these actors have starred in the same movie at least once
movies_both_actors <- intersect(
  dt.movie.actor[actor == top_actor_all$actor, unique(movie)],
  dt.movie.actor[actor == top_year_actor$actor, unique(movie)]
)
print(paste("Movie(s) with both top actors:", paste(movies_both_actors, collapse = ", ")))
```

### Question 7
```{r}
dt_decade <- dt.movie.actor[year >= 2010 & year <= 2019]

threshold <- quantile(dt_decade$votes, 0.90, na.rm = TRUE)

dt_popular <- dt_decade[votes >= threshold]

# PART A
top_10_actors <- dt_popular[, .(n_movies = .N), by = actor][order(-n_movies)][1:10]
print("Top 10 actors with the most popular movies (2010-2019):")
print(top_10_actors)

# PART B
top_year_participation <- dt_popular[, .(n_parts = .N), by = .(year, actor)][order(-n_parts)][1]
print("Year and actor with the most participations:")
print(top_year_participation)
```

## Part 2: `ggplot2`
### Question 1
```{r}
dt.movies.per.year <- dt.movie.actor[, .(n_movies = uniqueN(movie)), by = year]

ggplot(dt.movies.per.year, aes(x = year, y = n_movies)) +
  geom_bar(stat = "identity")
```
Movie production grew steadily from the early 1900s, with a sharp acceleration from the 1990s onwards, reflecting the globalisation of film production and the rise of digital filmmaking. There is a noticeable dip around the 1930s-40s, which can likely be explained by the Great Depression and World War II, both of which severely impacted the film industry. After the war, production recovered and grew gradually through the 1960s-80s before exploding in the 1990s and 2000s. One strange observation is that the last bar (2019) appears slightly lower than 2017 and 2018, which is unexpected given the general upward trend. This is most likely because the dataset was collected in January 2020, meaning not all 2019 movies had been added to IMDb yet at the time of collection.

### Question 2
```{r}
dt.votes <- unique(dt.movie.actor[, .(movie, votes)])

ggplot(dt.votes, aes(x = votes)) +
  geom_histogram() +
  scale_x_log10()
```
The distribution of IMDb votes per movie is extremely right-skewed. The vast majority of movies in the dataset have received very few votes while a tiny fraction of well-known blockbusters have accumulated hundreds of thousands or even millions of votes. This makes intuitive sense, as most movies in the dataset are obscure or low-budget productions that very few people have watched or rated. The log scale makes the shape much clearer, revealing that the distribution peaks before 100 votes and then tapers off gradually towards the right.

### Question 3
```{r}
dt.actors.per.movie <- unique(dt.movie.actor[, .(movie, n_actors)])

ggplot(dt.actors.per.movie, aes(x = n_actors)) +
  geom_histogram(binwidth = 1)
```
The vast majority of movies have exactly 4 actors listed, which is likely a structural feature of how IMDb records principal cast. Very few movies have more than 8 or fewer than 3 actors.


## Part 3
### Question 1
```{r}
# Filter to actors from top 50 most popular movies of 2010-2019 decade
top.movies.2010.2019 <- dt.movie.actor[year >= 2010 & year <= 2019, 
    .N, by = list(movie, votes)
    ][order(-votes)][1:50, movie]

actors.in.top <- dt.movie.actor[movie %in% top.movies.2010.2019, 
                                unique(actor)]

# Create bipartite graph with actors and movies
edges <- dt.movie.actor[movie %in% top.movies.2010.2019, 
                       list(actor, movie)]

# Create bipartite graph
g.bipartite <- graph_from_data_frame(edges, directed = FALSE)

# Mark bipartite attribute
V(g.bipartite)$type <- !V(g.bipartite)$name %in% unique(edges$movie)


# Summary and number of movie participations
summary(g.bipartite)
ecount(g.bipartite) 
```
There are 200 movie participations.

### Question 2
```{r}
# Create vertex list with type attribute
all.actors <- dt.movie.actor[movie %in% top.movies.2010.2019, list(name = unique(actor), type = TRUE)]
all.movies <- dt.movie.actor[movie %in% top.movies.2010.2019, list(name = unique(movie), type = FALSE)]
all.vertices <- rbind(all.actors, all.movies)

# Build bipartite graph
g <- graph.data.frame(dt.movie.actor[movie %in% top.movies.2010.2019,  
                                     list(movie, actor)
                                     ], directed = FALSE, vertices = all.vertices)

# Project to movie space 
g.movies <- bipartite.projection(g)$proj1
summary(g.movies)

# Calculate degree centrality
V(g.movies)$degree <- degree(g.movies)

# Remove movies with no connections (degree == 0)
g.movies.connected <- induced.subgraph(g.movies, vids = V(g.movies)[V(g.movies)$degree > 0])
summary(g.movies.connected)

# Calculate all centrality measures
V(g.movies.connected)$degree      <- degree(g.movies.connected)
V(g.movies.connected)$closeness   <- closeness(g.movies.connected)
V(g.movies.connected)$betweenness <- betweenness(g.movies.connected)
V(g.movies.connected)$evcent      <- evcent(g.movies.connected)$vector

# Convert to data.table for analysis
dt.g.movies <- data.table(get.data.frame(g.movies.connected, "vertices"))
head(dt.g.movies[order(-betweenness)], 10)
```

### Question 3
```{r}
# Top 20 by each centrality measure
head(dt.g.movies[order(-degree)], 20)
head(dt.g.movies[order(-closeness)], 20)
head(dt.g.movies[order(-betweenness)], 20)
head(dt.g.movies[order(-evcent)], 20)

# Report exact betweenness of top-ranked movie
dt.g.movies[order(-betweenness)][1, .(name, betweenness = round(betweenness, 4))]
```
Interpretation of Centrality Measures:
Degree Centrality: 
The Avengers series dominates with the highest degree of centrality, with each film having 11 connections. This indicates that these films share the most actors with other movies in the top 50, making them highly connected hubs in the actor-sharing network.
Closeness Centrality: 
X-Men: First Class (2011) has the highest closeness centrality score (0.0556), followed by other films with large casts, such as Silver Linings Playbook (2012) and X-Men: Days of Future Past (2014). On average, these movies are closer to the other movies, suggesting that they occupy more central positions in the network's overall structure.
Betweenness Centrality: 
Shutter Island (2010) takes the top spot with a score of 144.4286, followed by Now You See Me (2013) with 84.4286 and The Revenant (2015) with 63.50. These films act as crucial bridges, connecting different clusters of films through shared actors. Shutter Island is especially important in connecting parts of the network that would otherwise be disconnected.
Eigenvector Centrality: 
The three Avengers films received perfect scores of 1.0, indicating that they are connected to other highly connected movies. Captain America: Civil War (2016) and The Avengers (2012) also received high scores (0.78), demonstrating their connection to significant nodes in the network.

Key Insight: Different films are central for different reasons. The Avengers films are highly connected and influential, while Shutter Island acts as a critical bridge between different movie clusters. Films with high closeness centrality, such as X-Men: First Class, occupy more balanced positions within the network.

The exact betweenness centrality of the top-ranked movie is: Shutter Island (2010): 144.4286

### Question 4
```{r}
avg.clustering <- transitivity(g.movies.connected, type = "global")

print(paste("Average clustering coefficient:", round(avg.clustering, 4)))
```
The average clustering coefficient for the movies network is 0.6327. This value indicates a high degree of connectivity among movies through shared actors, reflecting the collaborative nature of the industry. However, it is slightly lower than the typical range of 0.75-0.85 usually observed in such networks. This deviation might be due to the specific selection of the top 50 most popular movies, which likely belong to diverse genres or franchises that do not overlap as much as a broader dataset might.

### Question 5
```{r}
# Define the target movie
my.movie <- "The Avengers (2012)"
print(paste("Selected Movie:", my.movie))

# Extract the ego graph (neighborhood) for the selected movie
g.movie.neighborhood <- make_ego_graph(
  g.movies.connected, 
  order = 1, 
  nodes = my.movie
)[[1]]

# Plot the neighborhood network
plot(
  g.movie.neighborhood, 
  vertex.size = ifelse(V(g.movie.neighborhood)$name == my.movie, 15, 8),
  vertex.color = ifelse(V(g.movie.neighborhood)$name == my.movie, "tomato", "lightblue"),
  vertex.label.cex = 0.7,
  vertex.label.color = "black",
  main = paste("Neighborhood of", my.movie)
)

# Calculate local clustering coefficient
local.cc <- transitivity(g.movies.connected, type = "local", vids = my.movie)
print(paste("Local Clustering Coefficient:", round(local.cc, 4)))

# Compute ratio to network average
cc.ratio <- local.cc / avg.clustering
print(paste("Ratio (Local / Network Average):", round(cc.ratio, 2)))

# Filter actors belonging to neighbors but excluding target actors
target.actors <- dt.movie.actor[movie == my.movie, actor]
neighbor.movies <- V(g.movie.neighborhood)$name[V(g.movie.neighborhood)$name != my.movie]

neighbor.actors <- dt.movie.actor[
  movie %in% neighbor.movies & !(actor %in% target.actors), 
  actor
]

# Find the most frequent actor
top.neighbor.actor <- names(sort(table(neighbor.actors), decreasing = TRUE)[1])
actor.freq <- sort(table(neighbor.actors), decreasing = TRUE)[1]

print(paste(
  "Top actor in neighborhood (but not in movie):", 
  top.neighbor.actor, 
  "with", 
  actor.freq, 
  "participations."
))
```
We chose to analyze the movie The Avengers (2012). The local clustering coefficient of this movie is 0.8571. The ratio of this movie's clustering coefficient to the network average (0.6327) is 1.35, indicating that this movie is part of a much more densely interconnected local neighborhood than the network average.

After analyzing the neighborhood, the actor with the most participations among these neighboring movies, but not having participated in the movie itself (according to the dataset's specific entries), is Chris Hemsworth (b.1983) with 3 participations.

### Question 6
```{r}
# Calculate empirical degree distribution
emp.deg.dist <- degree_distribution(g.movies.connected)

# Generate an equivalent random graph (Erdos-Renyi)
n.v <- vcount(g.movies.connected)
n.e <- ecount(g.movies.connected)
g.random <- erdos.renyi.game(n.v, n.e, type = "gnm")

# Calculate random graph degree distribution
rand.deg.dist <- degree_distribution(g.random)

# Plot distributions
max.deg <- max(length(emp.deg.dist), length(rand.deg.dist)) - 1

plot(
  0:(length(emp.deg.dist) - 1), 
  emp.deg.dist, 
  type = "b", 
  col = "blue", 
  pch = 16, 
  xlab = "Degree", 
  ylab = "Frequency / Probability", 
  main = "Degree Distribution: Movies vs Random Graph",
  xlim = c(0, max.deg)
)

points(
  0:(length(rand.deg.dist) - 1), 
  rand.deg.dist, 
  type = "b", 
  col = "red", 
  pch = 17
)

legend(
  "topright", 
  legend = c("Movies Network", "Random Graph"), 
  col = c("blue", "red"), 
  pch = c(16, 17), 
  lty = 1
)

# Identify bridge movie: high betweenness but not top degree
top5.betw <- head(dt.g.movies[order(-betweenness)]$name, 5)
top10.degree <- head(dt.g.movies[order(-degree)]$name, 10)

# Find the anomaly using set difference
bridge.movie.list <- setdiff(top5.betw, top10.degree)
bridge.movie <- bridge.movie.list[1]
bridge.betweenness <- dt.g.movies[name == bridge.movie, betweenness]

print(paste("Bridge Movie:", bridge.movie))
print(paste("Exact Betweenness:", round(bridge.betweenness, 4)))
```
Comparing the degree distribution of our movies network with a random (Erdős-Rényi) graph reveals significant differences. While the random graph's degree distribution is bell-shaped and concentrated around its mean, our network shows a right-skewed distribution with a longer tail. This reflects the "hub" phenomenon typical of the film industry, where highly successful franchises create nodes with an exceptionally high number of connections compared to an average movie.

A movie acting as a critical bridge in this network is Now You See Me (2013). It ranks in the top 5 by betweenness centrality but does not appear in the top 10 by degree centrality. Its exact betweenness centrality is 84.4286. This movie serves as a crucial bridge because it acts as the shortest path connecting distinct, otherwise isolated clusters (such as different actor circles or genres), despite having relatively few direct connections itself.

